Analysis of the time-dependent response to Estradiol through ChIP-Seq data and the VULCAN package

### 2018-11-26 S2_ChIP-seq_ER/qrime/03.QRIMESceneInvestigatonCombined.Rmd
remove big files

---
title: "Supplementary"
author: "Andrew Holding & Matt Eldridge"
date: "9 January 2017"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    fig_caption: yes
header-includes:
  - \usepackage{longtable}
  - \usepackage{booktabs}
---

@@

# Introduction

This report is from an analysis of the RIME-TMT proteomics data for an experiment carried out by Andrew Holding to quantify specific interactors of estrogen receptor (ER) and forkhead box protein A1 (FOXA1) in MCF-7 cells.

The RIME (Rapid Immunoprecipitation of Endogenous Proteins) technique was used to pull down ER and FOXA1 and their interacting proteins, which were subsequently assessed and quantitated using mass spectrometry with tandem mass tags (TMT). TMT runs were carried out for 3 biological replicates, estrogen-starved MCF-7 cells from the same cell line at different times/passages, before and after addition of estrogen with samples taken at 3 different time points -- 0, 45 and 90 minutes after estrogen addition. A pull down of Immunoglobulin G (IgG) was also run as a control.

The digests from two of the biological replicates (PR622 and PR650) were analyzed with 3 separate runs of the mass spectrometer to increase coverage; data from the repeated runs were combined prior to the analysis carried out here. The other sample (PR590) was run once only.

`r figr("samples_table", TRUE, type="Table")` shows the isobaric tags used for each sample within each of the runs.

```{r samples_table, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
print(
  xtable(
    metadata %>% select(Run, Tag, Group) %>% spread(Run, Tag),
    caption = "Isobaric tags used for each sample (group) and run."
  ),
  include.rownames = FALSE,
  size = "\\setlength{\\tabcolsep}{12pt}"
)
```

@@

```{r normalized_intensity_plots, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Density plots of normalized intensities from each TMT run where scale normalization was applied to peptide intensities that include misssing values.", fig.height=8, fig.width=5}

analysis <- "IncludeMissingValues_ScaleNormalization"

@@

```{r pca_plot_er, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Principal Componenent Analysis for ER-interactng proteins sampled in all 3 runs. The first two principal components are displayed.", fig.height=5}

metadata <- metadata %>% filter(grepl("^ER", Group))

@@

The IgG control was used to assess non-specific binding. The binding of a protein is considered to be non-specific if the log~2~ fold change relative to the IgG control is less than `r controlLogFoldChangeThreshold`. In the differential binding analysis, protein intensity values are fitted to a linear model and the fitted values for each condition being compared are also compared to the IgG control. For each comparison of two groups, the maximum log~2~ fold change from each of the two groups above the IgG control is used to determine if the binding is specific.

```{r contrast_1, include = FALSE}
contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")
processingDescription <- "excluding peptides with missing intensities, no normalization"
analysis <- "ExcludeMissingValues_NoNormalization"
```

```{r results_1, include = FALSE}

```{r ma_plot_1, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("MA plot of the average intensity against log~2~ fold change for the ", contrastName, " comparison (", processingDescription, "). Top ranking differentially-expressed proteins with false discovery rate below 0.05 are highlighted in pink. Open circles indicate that the protein is non-specific from the IgG control comparison.", sep = ""), fig.height=6}
print(maPlot(results,
             significanceLevel = significanceLevel,
             controlLogFoldChangeThreshold = controlLogFoldChangeThreshold,
             minLogFoldChangeForLabelling = 1.5,
             maxNumberLabelledProteins = 20,
             pointSize = 1.5))
```

```{r volcano_plot_1, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Volcano plot of the average intensity against log~2~ fold change for the ", contrastName, " comparison (", processingDescription, "). Top ranking differentially-expressed proteins with false discovery rate below 0.05 are highlighted in pink. Open circles indicate that the protein is non-specific from the IgG control comparison.", sep = ""), fig.height=6}
print(volcanoPlot(results,
                  significanceLevel = significanceLevel,
                  minLogFoldChange = -2.0,
                  maxLogFoldChange = 2.75,
                  controlLogFoldChangeThreshold = controlLogFoldChangeThreshold,
                  minLogFoldChangeForLabelling = 1.5,
                  maxNumberLabelledProteins = 20,
                  pointSize = 1.5))
```

```{r pvalue_histogram_1, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Histogram of p-values for the ", contrastName, " comparison (", processingDescription, ")", sep = ""), fig.height=3.5, fig.width=4.5}
print(histogram(results, "P.Value", xlab = "p-value"))
```

```{r qq_plot_1, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("QQ plot of the adjusted p-values for the ", contrastName, " comparison (", processingDescription, ")", sep = ""), fig.height=5, fig.width=5}
q <- results %>%
  select(adj.P.Val) %>%
  filter(!is.na(adj.P.Val)) %>%
  arrange(adj.P.Val) %>%
  unlist %>%
  as.numeric
qqplot(-log10(ppoints(length(q))),
       -log10(q),
       xlab=expression(Expected~~-log[10](italic(q))),
       ylab=expression(Observed~~-log[10](italic(q))))
abline(0, 1, col = "red")
```

@@

```{r contrast_3, include = FALSE}
contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")
processingDescription <- "excluding peptides with missing intensities, quantile normalization excluding IgG control"
analysis <- "ExcludeMissingValues_QuantileNormalizationExcludingIgG"
```

@@

@@

```{r contrast_4, include = FALSE}
contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")
processingDescription <- "excluding peptides with missing intensities, scale normalization"
analysis <- "ExcludeMissingValues_ScaleNormalization"
```

@@

```{r volcano_plot_4, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Volcano plot of the average intensity against log~2~ fold change for the ", contrastName, " comparison (", processingDescription, "). Top ranking differentially-expressed proteins with false discovery rate below 0.05 are highlighted in pink. Open circles indicate that the protein is non-specific from the IgG control comparison.", sep = ""), fig.height=6}
print(volcanoPlot(results,
                  significanceLevel = significanceLevel,
                  minLogFoldChange = -2.0,
                  maxLogFoldChange = 2.75,
                  controlLogFoldChangeThreshold = controlLogFoldChangeThreshold,
                  minLogFoldChangeForLabelling = 1.5,
                  maxNumberLabelledProteins = 50,
                  pointSize = 1.5))
```

```{r pvalue_histogram_4, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Histogram of p-values for the ", contrastName, " comparison (", processingDescription, ")", sep = ""), fig.height=3.5, fig.width=4.5}
print(histogram(results, "P.Value", xlab = "p-value"))
```

```{r qq_plot_4, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("QQ plot of the adjusted p-values for the ", contrastName, " comparison (", processingDescription, ")", sep = ""), fig.height=5, fig.width=5}
q <- results %>%
  select(adj.P.Val) %>%
  filter(!is.na(adj.P.Val)) %>%
  arrange(adj.P.Val) %>%
  unlist %>%
  as.numeric
qqplot(-log10(ppoints(length(q))),
       -log10(q),
       xlab=expression(Expected~~-log[10](italic(q))),
       ylab=expression(Observed~~-log[10](italic(q))))
abline(0, 1, col = "red")
```

@@

```{r contrast_5, include = FALSE}
contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")
processingDescription <- paste("excluding peptides with missing intensities, scale normalization using top", numberOfPeptidesWithHighestIgG, "IgG peptides")
analysis <- "ExcludeMissingValues_ScaleNormalizationTopIgG"
```

@@

```{r volcano_plot_5, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Volcano plot of the average intensity against log~2~ fold change for the ", contrastName, " comparison (", processingDescription, "). Top ranking differentially-expressed proteins with false discovery rate below 0.05 are highlighted in pink. Open circles indicate that the protein is non-specific from the IgG control comparison.", sep = ""), fig.height=6}
print(volcanoPlot(results,
                  significanceLevel = significanceLevel,
                  minLogFoldChange = -2.0,
                  maxLogFoldChange = 2.75,
                  controlLogFoldChangeThreshold = controlLogFoldChangeThreshold,
                  minLogFoldChangeForLabelling = 1.5,
                  maxNumberLabelledProteins = 50,
                  pointSize = 1.5))
```

```{r pvalue_histogram_5, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Histogram of p-values for the ", contrastName, " comparison (", processingDescription, ")", sep = ""), fig.height=3.5, fig.width=4.5}
print(histogram(results, "P.Value", xlab = "p-value"))
```

```{r qq_plot_5, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("QQ plot of the adjusted p-values for the ", contrastName, " comparison (", processingDescription, ")", sep = ""), fig.height=5, fig.width=5}
q <- results %>%
  select(adj.P.Val) %>%
  filter(!is.na(adj.P.Val)) %>%
  arrange(adj.P.Val) %>%
  unlist %>%
  as.numeric
qqplot(-log10(ppoints(length(q))),
       -log10(q),
       xlab=expression(Expected~~-log[10](italic(q))),
       ylab=expression(Observed~~-log[10](italic(q))))
abline(0, 1, col = "red")
```

```{r contrast_6, include = FALSE}
contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")
processingDescription <- "KNN nearest neighbour averaging imputation of missing values, quantile normalization"
analysis <- "ImputeKNN_QuantileNormalization"
```

```{r results_6, include = FALSE}

```{r ma_plot_6, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("MA plot of the average intensity against log~2~ fold change for the ", contrastName, " comparison (", processingDescription, "). Top ranking differentially-expressed proteins with false discovery rate below 0.05 are highlighted in pink. Open circles indicate that the protein is non-specific from the IgG control comparison.", sep = ""), fig.height=6}
print(maPlot(results,
             significanceLevel = significanceLevel,
             controlLogFoldChangeThreshold = controlLogFoldChangeThreshold,
             minLogFoldChangeForLabelling = 1.5,
             maxNumberLabelledProteins = 50,
             pointSize = 1.5))
```

```{r volcano_plot_6, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Volcano plot of the average intensity against log~2~ fold change for the ", contrastName, " comparison (", processingDescription, "). Top ranking differentially-expressed proteins with false discovery rate below 0.05 are highlighted in pink. Open circles indicate that the protein is non-specific from the IgG control comparison.", sep = ""), fig.height=6}
print(volcanoPlot(results,
                  significanceLevel = significanceLevel,
                  minLogFoldChange = -2.0,
                  maxLogFoldChange = 2.75,
                  controlLogFoldChangeThreshold = controlLogFoldChangeThreshold,
                  minLogFoldChangeForLabelling = 1.5,
                  maxNumberLabelledProteins = 50,
                  pointSize = 1.5))
```

```{r pvalue_histogram_6, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Histogram of p-values for the ", contrastName, " comparison (", processingDescription, ")", sep = ""), fig.height=3.5, fig.width=4.5}
print(histogram(results, "P.Value", xlab = "p-value"))
```

```{r qq_plot_6, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("QQ plot of the adjusted p-values for the ", contrastName, " comparison (", processingDescription, ")", sep = ""), fig.height=5, fig.width=5}
q <- results %>%
  select(adj.P.Val) %>%
  filter(!is.na(adj.P.Val)) %>%
  arrange(adj.P.Val) %>%
  unlist %>%
  as.numeric
qqplot(-log10(ppoints(length(q))),
       -log10(q),
       xlab=expression(Expected~~-log[10](italic(q))),
       ylab=expression(Observed~~-log[10](italic(q))))
abline(0, 1, col = "red")
```

# Comparison of differential binding analysis approaches

In this section, comparisons of the differential binding results obtained using differing methods for handling missing values and varying normalization techniques are presented. The B-statistic (log odds) is used to rank the proteins from most significantly differentially expressed to least significant. Scatter plots of both the B-statistic generated for the two methods being compared and the rank are presented. The protein with the highest B-statistic, i.e. the most significantly differentially expressed protein, is given a rank of 1.

Differing imputation methods have relatively little effect on the results for most proteins. This is largely because only a small proportion of the proteins have peptide-level measurements containing missing values. There are a few proteins that are only represented in the differential binding results when imputation is carried out, for which all peptide-level measurements contain missing values. These proteins are not represented in the comparions with the analysis in which measurements with missing values are removed.

Different normalization approaches have a more substantial impact on the differential binding analysis.

## Removal of missing values vs KNN imputation (quantile normalization in both)

```{r exclude_missing_vs_knn_imputation, include = FALSE}

contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")

analysis <- "ExcludeMissingValues_QuantileNormalization"
data <- combinedProteinData[[analysis]]
exprs(data) <- log2(exprs(data))
results1 <- createContrastResults(
  data,
  fittedLinearModels[[analysis]],
  fittedLinearModelsExcludingControls[[analysis]],
  fittedContrasts[[analysis]],
  runs,
  contrastGroups,
  controlGroup,
  proteinInfo)

analysis <- "ImputeKNN_QuantileNormalization"
data <- combinedProteinData[[analysis]]
exprs(data) <- log2(exprs(data))
results2 <- createContrastResults(
  data,
  fittedLinearModels[[analysis]],
  fittedLinearModelsExcludingControls[[analysis]],
  fittedContrasts[[analysis]],
  runs,
  contrastGroups,
  controlGroup,
  proteinInfo)

results <- inner_join(
  results1 %>% select(Protein, B1 = B),
  results2 %>% select(Protein, B2 = B),
  by = "Protein"
  ) %>%
  filter(complete.cases(.)) %>%
  mutate(rank1 = rank(-B1), rank2 = rank(-B2))
```

```{r exclude_missing_vs_knn_imputation_B_plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Plot of B-statistic (log odds) for differential expression of proteins for the ", contrastName, " contrast, comparing results following exclusion of missing values and KNN imputation.", sep = ""), fig.height=2.9, fig.width=3.5}
print(scatterPlot(results, "B1", "B2", "B-statistic (excluding missing values)", "B-statistic (KNN imputation)"))
```

```{r exclude_missing_vs_knn_imputation_rank_plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Plot of the rank of differentially-expressed proteins for the ", contrastName, " contrast, comparing results following exclusion of missing values and KNN imputation.", sep = ""), fig.height=2.9, fig.width=3.5}
print(scatterPlot(results, "rank1", "rank2", "Rank (excluding missing values)", "Rank (KNN imputation)"))
```

## Removal of missing values vs lowest value imputation (quantile normalization in both)

```{r exclude_missing_vs_min_imputation, include = FALSE}

contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")

analysis <- "ExcludeMissingValues_QuantileNormalization"
data <- combinedProteinData[[analysis]]
exprs(data) <- log2(exprs(data))
results1 <- createContrastResults(
  data,
  fittedLinearModels[[analysis]],
  fittedLinearModelsExcludingControls[[analysis]],
  fittedContrasts[[analysis]],
  runs,
  contrastGroups,
  controlGroup,
  proteinInfo)

analysis <- "ImputeMin_QuantileNormalization"
data <- combinedProteinData[[analysis]]
exprs(data) <- log2(exprs(data))
results2 <- createContrastResults(
  data,
  fittedLinearModels[[analysis]],
  fittedLinearModelsExcludingControls[[analysis]],
  fittedContrasts[[analysis]],
  runs,
  contrastGroups,
  controlGroup,
  proteinInfo)

results <- inner_join(
  results1 %>% select(Protein, B1 = B),
  results2 %>% select(Protein, B2 = B),
  by = "Protein"
  ) %>%
  filter(complete.cases(.)) %>%
  mutate(rank1 = rank(-B1), rank2 = rank(-B2))
```

```{r exclude_missing_vs_min_imputation_B_plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Plot of B-statistic (log odds) for differential expression of proteins for the ", contrastName, " contrast, comparing results following exclusion of missing values and lowest value imputation.", sep = ""), fig.height=2.9, fig.width=3.5}
print(scatterPlot(results, "B1", "B2", "B-statistic (excluding missing values)", "B-statistic (lowest value imputation)"))
```

```{r exclude_missing_vs_min_imputation_rank_plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Plot of the rank of differentially-expressed proteins for the ", contrastName, " contrast, comparing results following exclusion of missing values and lowest value imputation.", sep = ""), fig.height=2.9, fig.width=3.5}
print(scatterPlot(results, "rank1", "rank2", "Rank (excluding missing values)", "Rank (lowest value imputation)"))
```

## Quantile vs Scale normalization (missing values excluded in both)

```{r quantile_vs_scale_normalization, include = FALSE}

contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")

analysis <- "ExcludeMissingValues_QuantileNormalization"
data <- combinedProteinData[[analysis]]
exprs(data) <- log2(exprs(data))
results1 <- createContrastResults(
  data,
  fittedLinearModels[[analysis]],
  fittedLinearModelsExcludingControls[[analysis]],
  fittedContrasts[[analysis]],
  runs,
  contrastGroups,
  controlGroup,
  proteinInfo)

analysis <- "ExcludeMissingValues_ScaleNormalization"
data <- combinedProteinData[[analysis]]
exprs(data) <- log2(exprs(data))
results2 <- createContrastResults(
  data,
  fittedLinearModels[[analysis]],
  fittedLinearModelsExcludingControls[[analysis]],
  fittedContrasts[[analysis]],
  runs,
  contrastGroups,
  controlGroup,
  proteinInfo)

results <- inner_join(
  results1 %>% select(Protein, B1 = B),
  results2 %>% select(Protein, B2 = B),
  by = "Protein"
  ) %>%
  filter(complete.cases(.)) %>%
  mutate(rank1 = rank(-B1), rank2 = rank(-B2))
```

```{r scale_vs_quantile_normalization_B_plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Plot of B-statistic (log odds) for differential expression of proteins for the ", contrastName, " contrast, comparing results following quantile and scale normalization.", sep = ""), fig.height=2.9, fig.width=3.5}
print(scatterPlot(results, "B1", "B2", "B-statistic (quantile normalization)", "B-statistic (scale normalization)"))
```

```{r scale_vs_quantile_normalization_rank_plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Plot of the rank of differentially-expressed proteins for the ", contrastName, " contrast, comparing results following quantile and scale normalization.", sep = ""), fig.height=2.9, fig.width=3.5}
print(scatterPlot(results, "rank1", "rank2", "Rank (quantile normalization)", "Rank (scale normalization)"))
```

## Scale normalization based on all peptides vs peptides with highest IgG intensities (missing values excluded in both)

```{r scale_normalization_all_vs_topIgG, include = FALSE}

contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")

analysis <- "ExcludeMissingValues_ScaleNormalization"
data <- combinedProteinData[[analysis]]
exprs(data) <- log2(exprs(data))
results1 <- createContrastResults(
  data,
  fittedLinearModels[[analysis]],
  fittedLinearModelsExcludingControls[[analysis]],
  fittedContrasts[[analysis]],
  runs,
  contrastGroups,
  controlGroup,
  proteinInfo)

analysis <- "ExcludeMissingValues_ScaleNormalizationTopIgG"
data <- combinedProteinData[[analysis]]
exprs(data) <- log2(exprs(data))
results2 <- createContrastResults(
  data,
  fittedLinearModels[[analysis]],
  fittedLinearModelsExcludingControls[[analysis]],
  fittedContrasts[[analysis]],
  runs,
  contrastGroups,
  controlGroup,
  proteinInfo)

results <- inner_join(
  results1 %>% select(Protein, B1 = B),
  results2 %>% select(Protein, B2 = B),
  by = "Protein"
  ) %>%
  filter(complete.cases(.)) %>%
  mutate(rank1 = rank(-B1), rank2 = rank(-B2))
```

```{r scale_normalization_all_vs_topIgG_B_plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Plot of B-statistic (log odds) for differential expression of proteins for the ", contrastName, " contrast, comparing results following scale normalization based on all peptides and those with the highest IgG intensities.", sep = ""), fig.height=2.5, fig.width=3.5}
print(scatterPlot(results, "B1", "B2", "B-statistic (scale normalization based on all peptides)", "B-statistic (scale normalization, peptides with highest IgG)"))
```

```{r scale_normalization_all_vs_topIgG_rank_plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=paste("Plot of the rank of differentially-expressed proteins for the ", contrastName, " contrast, comparing results following scale normalization based on all peptides and those with the highest IgG intensities.", sep = ""), fig.height=2.5, fig.width=3.5}
print(scatterPlot(results, "rank1", "rank2", "Rank (scale normalization based on all peptides)", "Rank (scale normalization, peptides with highest IgG)"))
```

# Comparisoin of variance between specfic and non-specific binding

```{r  echo=FALSE, warning=FALSE, message=FALSE}

maPlot <- function(differentialExpressionResults, selectedGenes = NULL,
                   xlab = expression(average~log[2](intensity)),
                   ylab = expression(log[2](fold~change)),
                   significanceLevel = 0.05,
                   minLogFoldChangeForLabelling = 1.5,
                   maxNumberLabelledProteins = 50,
                   controlLogFoldChangeThreshold = -Inf,
                   pointSize = 2.5) {
  if (!"logFCcontrol" %in% colnames(differentialExpressionResults))
    differentialExpressionResults$logFCcontrol <- Inf

  differentialExpressionResults$group <-
    ifelse(differentialExpressionResults$Gene %in% selectedGenes, 0,
    ifelse(!is.na(differentialExpressionResults$adj.P.Val) & differentialExpressionResults$adj.P.Val <= significanceLevel & differentialExpressionResults$logFCcontrol >= controlLogFoldChangeThreshold, 1,
    ifelse(!is.na(differentialExpressionResults$adj.P.Val) & differentialExpressionResults$adj.P.Val <= significanceLevel, 2,
    ifelse(abs(differentialExpressionResults$logFC) >= minLogFoldChangeForLabelling & differentialExpressionResults$logFCcontrol >= controlLogFoldChangeThreshold, 3,
    ifelse(abs(differentialExpressionResults$logFC) >= minLogFoldChangeForLabelling, 4,
    ifelse(differentialExpressionResults$logFCcontrol >= controlLogFoldChangeThreshold, 5, 6))))))

  # 0 - selected
  # 1 - significant, specific
  # 2 - significant, non-specific
  # 3 - non-significant, large logfc, specific
  # 4 - non-significant, large logfc, non-specific
  # 5 - specific
  # 6 - non-specific

  differentialExpressionResults <- differentialExpressionResults %>% arrange(desc(group))

  differentialExpressionResults$group <- factor(differentialExpressionResults$group, 0:6)

  differentialExpressionResults <- differentialExpressionResults %>%
    mutate(label = paste(" ", Gene)) %>%
    mutate(logFCcontrolAlpha = pmin(logFCcontrol, 2.0))

  labelSubset <- bind_rows(
    differentialExpressionResults %>%
      filter(group != 0 & adj.P.Val <= significanceLevel) %>%
      arrange(desc(abs(logFC))) %>%
      slice(1:(maxNumberLabelledProteins / 2)),
    differentialExpressionResults %>%
      filter(group != 0 & abs(logFC) >= minLogFoldChangeForLabelling) %>%
      arrange(desc(abs(logFC))) %>%
      slice(1:maxNumberLabelledProteins)
  ) %>%
    unique %>%
    slice(1:maxNumberLabelledProteins)

    if (maxNumberLabelledProteins == 0) labelSubset <- labelSubset %>% slice(0)

  # plot <- ggplot(differentialExpressionResults, aes(x = AveExpr, y = logFC, colour = group, size = group, shape = group, alpha = logFCcontrolAlpha))
  plot <- ggplot(differentialExpressionResults, aes(x = AveExpr, y = logFC, colour = group))+ coord_cartesian(ylim = c(-3.5,3.5))
  plot <- plot + geom_hline(yintercept = 0, color="gray50", size = 0.5)
  plot <- plot + geom_point(alpha=0.75)
  plot <- plot + geom_text(data = labelSubset, aes(label = label), hjust = 0, size = 3.5)
  plot <- plot + geom_point(data = subset(differentialExpressionResults, group == 0), alpha = 0.7)
  plot <- plot + geom_text(data = subset(differentialExpressionResults, group == 0), aes(label = label), hjust = 0, size = 4.0, alpha = 1)
  plot <- plot + scale_colour_manual(values = c("green", "red", "blue", "red", "blue", "red", "blue"), drop = FALSE)
  # plot <- plot + scale_colour_manual(values = c("blue", "deeppink3", "deeppink3", "gray50", "gray50", "gray50", "gray50"), drop = FALSE)
  # plot <- plot + scale_size_manual(values = c(1.2 * pointSize, pointSize, pointSize, 0.8 * pointSize, 0.8 * pointSize, 0.6 * pointSize, 0.6 * pointSize), drop = FALSE)
  # plot <- plot + scale_shape_manual(values = c(16, 16, 1, 16, 1, 16, 1), drop = FALSE)
  #plot <- plot + scale_alpha(range = c(0.2, 1.0))
  plot <- plot + xlab(xlab)
  plot <- plot + ylab(ylab)
  plot <- plot + theme_bw()
  plot <- plot + theme(
    text = element_text(size = 14),
    legend.position = "none"
  )

  save(differentialExpressionResults,file="Results/QRIME.RData") 
  return(plot)
}

contrastGroups <- c("ER 45min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")

analysis <- "ExcludeMissingValues_QuantileNormalization"

results <- createContrastResults(
  data,
  fittedLinearModels[[analysis]],
  fittedLinearModelsExcludingControls[[analysis]],
  fittedContrasts[[analysis]],
  runs,
  c("ER 45min", "ER 0min"),
  controlGroup,
  proteinInfo)

```{r MA_specificVsNonSpecfic, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="MA plot showing the distribution of non-specific binding. Non-specific binding was identified as proteins that showed less than a log2 enrichment over the IgG channel at either 0 or 45 minutes. The majority of non-specific binding (blue) is found to show less change between time points whereas the specific interactions (red) shows a broader distribution.", fig.height=2.5, fig.width=3.5}

print(maPlot(results,
             significanceLevel = 0.5,
             controlLogFoldChangeThreshold = 1.0,
             minLogFoldChangeForLabelling = 1.5,
             maxNumberLabelledProteins = 0,
             pointSize = 1.5))
```

```{r  echo=FALSE, warning=FALSE, message=FALSE}

load("Results/QRIME.RData")
as.numeric(differentialExpressionResults$group)%%2->differentialExpressionResults$group

tmp<-differentialExpressionResults

```{r density_specificVsNonSpecfici, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Density plot right showing the distribution of non-specific binding. Non-specific binding was identified as proteins that showed less than a log2 enrichment over the IgG channel at either 0 or 45 minutes. The majority of non-specific binding (blue) is found to show less change between time points whereas the specific interactions (red) shows a broader distribution.", fig.height=3.5, fig.width=4.5}

plot(density(tmp$logFC[tmp$group==1]),col="blue",ylab="Density", xlab="Fold Change", main="", xlim=c(3,-3), lwd=2)
lines(density(tmp$logFC[tmp$group==0]),col="red",lwd=2)

```{r box_specificVsNonSpecfici, echo=FALSE, warning=FALSE, message=FALSE, fig.cap=" Box plot showing the variance of non-specific vs non-specific binding. Non-specific binding was identified as proteins that showed less than a log2 enrichment over the IgG channel at either 0 or 45 minutes. The the variance of the log fold changes for the non-specific binding interactions was significantly less than that of the specific interactions (F-test, one-sided, p-value < 2.2e*-16).", fig.height=3.5, fig.width=4.5}

boxplot(tmp$logFC[tmp$group==1],tmp$logFC[tmp$group==0],ylab="Fold Change", names=c("Non-specific","Specific"), col=c("#0000FFAA","#FF0000AA"))

```{r  echo=FALSE, warning=FALSE, message=FALSE}

var.test(tmp$logFC[tmp$group==1],tmp$logFC[tmp$group==0], alternative="less")

@@

The following tables contain the top ranking differentially expressed proteins for each comparison. Included are all proteins that reach a statistical signficance of `r significanceLevel` in terms of the adjusted p-value and those with an absolute log~2~ fold change of `r logFoldChangeThreshold` or above.

@@

  print(
    xtable(
      resultsTable,
      align = "lllrrrrrrl",
      display = c("d", "s", "s", "d", "f", "f", "g", "f", "f", "s"),
      caption = paste("Top ranking differentially expressed proteins from the", contrastName, "comparison,  sorted by log2 fold change.")
    ),
    include.rownames = FALSE,
    tabular.environment = "longtable",
    floating = FALSE,
    add.to.row = addtorow
  )
}
```

>>>>

---
title: "QRIME Scene Investigation"
author: "Andrew Holding & Matt Eldridge"
date: "9 January 2017"
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
  html_document:
    toc: yes
header-includes:
- \usepackage{longtable}
- \usepackage{booktabs}
---

@@

# SampleSheet

```{r samples_table, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
print(
  xtable(
    metadata %>% select(Run, Tag, Group) %>% spread(Run, Tag),
    caption = "Isobaric tags used for each sample (group) and run."
  ),
  include.rownames = FALSE,
  size = "\\setlength{\\tabcolsep}{12pt}"
)

print(
  xtable(
    contrastsTable %>% select(Contrast),
    caption = "Configured Constrasts."
  ),
  include.rownames = FALSE,
  size = "\\setlength{\\tabcolsep}{12pt}"
)
```

@@

```{r normalized_intensity_plots, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Density plots of normalized intensities from each TMT run where QUANTILE normalization was applied to peptide intensities that include misssing values.", fig.height=6.5, fig.width=5}

analysis <- "IncludeMissingValues_QuantileNormalization"

@@

@@

@@

```{r contrast_3, include = FALSE}
contrastGroups <- c("ER 90min", "ER 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")
processingDescription <- "excluding peptides with missing intensities, quantile normalization"
analysis <- "ExcludeMissingValues_QuantileNormalization"
```

@@

print(contrastGroups)

@@

```{r contrast_4, include = FALSE}
contrastGroups <- c("FOXA1 45min", "FOXA1 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")
processingDescription <- "excluding peptides with missing intensities, quantile normalization"
analysis <- "ExcludeMissingValues_QuantileNormalization"
```

@@

@@

```{r contrast_5, include = FALSE}
contrastGroups <- c("FOXA1 90min", "FOXA1 0min")
contrastName <- paste(contrastGroups, collapse = " vs ")
processingDescription <- "excluding peptides with missing intensities, quantile normalization"
analysis <- "ExcludeMissingValues_QuantileNormalization"
```

@@

@@

The following tables contain the top ranking differentially bound proteins for each comparison. Included are all proteins that reach a statistical signficance of `r significanceLevel` in terms of the adjusted p-value and those with an absolute log~2~ fold change of `r logFoldChangeThreshold` or above.

@@

  print(
    xtable(
      resultsTable,
      align = "lllrrrrrrl",
      display = c("d", "s", "s", "d", "f", "f", "g", "f", "f", "s"),
      caption = paste("Top ranking differentially bound proteins from the", contrastName, "comparison,  sorted by log2 fold change.")
    ),
    include.rownames = FALSE,
    tabular.environment = "longtable",
    floating = FALSE,
    add.to.row = addtorow
  )
}
```

### 2018-11-26 S2_ChIP-seq_ER/rmarkdown.Rmd
remove big files

### VULCAN and qPLEX-RIME
We set to testing the performance of VULCAN against a complementary experimental approach called qPLEX-RIME (Holding et al., submitted), which aims at identifyin interactors of ER within the ER-chromatin complex at estrogen-responding promoters. We have QRIME results for MCF7 cells treated with estradiol at both 45 and 90 minutes, the same experimental setup for the VULCAN input dataset.
```{r qrime, message=FALSE, warning=FALSE}
# Load imported vulcan object
load("results/001_vobj.rda")

>>>>

### VULCAN and qPLEX-RIME
We set to testing the performance of VULCAN against a complementary experimental approach called qPLEX-RIME (Papachristou et al., under review), which aims at identifyin interactors of ER within the ER-chromatin complex at estrogen-responding promoters. We have QRIME results for MCF7 cells treated with estradiol at both 45 and 90 minutes, the same experimental setup for the VULCAN input dataset.
```{r qrime, message=FALSE, warning=FALSE}
# Load imported vulcan object
load("results/001_vobj.rda")
